---
title: "Setting Up and Running PHRAPL"
author: "Nathan Jackson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Setting Up and Running PHRAPL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This is a brief tutorial that will walk you through the necessary steps required for doing model selection with PHRAPL, using a toy dataset. The steps involved are as follows:

1. Instal PHRAPL and its dependencies
2. Prepare the dataset
3. Analyze the dataset
4. View results

This tutorial assumes you are using a Mac. For more detailed information about certain tasks, take a look at the help files within PHRAPL (for a list of these, type `library(help=phrapl)` once the package is installed).

The included files used in this tutorial are

1. trees.tre, which gives 10 sample trees for three simulated populations
2. cladeAssignments.txt, which gives the assignment of individuals to populations
3. InstallPhraplDependencies.R, which gives code for installing dependencies used by phrapl

# Installing PHRAPL and its dependencies

## 1. Install ms

The first step is to install Richard Hudson's coalescent simulator, ms, which PHRAPL uses to simulate trees under a model. To do this, first download the program available here:

    http://home.uchicago.edu/rhudson1/source/mksamples.html

Then, to extract the files, in the terminal, navagate to the ms directory and type

    tar -xvf ms.gz

To compile the program (per the ms manual), navagate to msdir (by typing `cd msdir`) and type

    gcc -o ms ms.c streec.c rand1.c -lm

If this doesn't work, try this instead:

    gcc -o ms ms.c streec.c rand2.c -lm

Then copy ms to local bin by typing

    cp ms /usr/local/bin

## 2. Install Xcode

Next, install the integrated development environment, Xcode, which can be downloaded from the following webpage:

    https://developer.apple.com/xcode/downloads/

##3. Install perl

The scripting language, perl, comes pre-intalled on Mac OS X systems. However, to make sure, type `perl -v`, which will give the version installed. If a new version needs to be installed for whatever reason, you can download it from the following webpage:

    http://www.perl.org/get.html

##4. Install PHRAPL from Github

PHRAPL can currently be found on Github. To install it, first install the R package, `devtools` by typing

    install.packages("devtools")
    
Then load devtools (`library(devtools)`) and type

    devtools::install_github("bomeara/phrapl")
  
##5. Install PHRAPL R package dependencies

PHRAPL uses functions from a variety of other R pacakges. To install these, open R and load the PHRAPL library by typing `library(phrapl)`. Then type

    source(paste(path.package("phrapl"),"/extdata/InstallPhraplDependencies.R",sep=""))

Now PHRAPL is ready to run. Type `library(help=phrapl)` to get a list of functions with documentation. To open a help file for a particular function, type `?function_name`.

#Preparing your dataset for a PHRAPL analysis









The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
